\documentclass{book}
\usepackage[utf8]{inputenc}


\usepackage{natbib}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{ascmac}
\usepackage{braket}
\usepackage{makeidx}
\makeindex
\usepackage{textcomp}
\usepackage{comment}
\usepackage{authblk}
\usepackage{xcolor}

\begin{document}
\title{Lecture Note \\ Quantum Mechanics of Light and Matters}
\author{Yasuyuki Ozeki}
\affil{Department of Electrical Engineering and Information Systems \\ The University of Tokyo}
\date{July 25, 2020}

\maketitle
\tableofcontents
\mainmatter


\input{Chap1_introduction}
\input{Chap2_noise_in_optical_measurement}
\input{Chap3_quantum-mechanical_harmonic_oscillator}


\chapter{Evolution of quantum states}
In the previous chapter, we saw quantum state of a harmonic oscillator in various representations. In quantum optics, we control the quantum state of light and investigate its properties by the measurement of observables. 

Here we introduce various operators that can be used to describe how we control the quantum state. Furthermore, we see that changing quantum state is equivalent to controlling the operator of observable. In this way, we can control the quantum state or the operator of observable. The former is called Schr\"odinger picture, while the latter is called Heisenberg picture. Schr\"odinger picture gives us intuitive picture of quantum state, while the latter allows us to simplify the calculation of observables. 


\section{Schor\"odinger picture}
In the Schor\"odinger picture, we consider that the quantum state changes according to the Schr\"odinger equation given by
\begin{equation}
  i\hbar \frac{d}{dr}\ket {\psi(r)} = \hat H \ket{\psi(r)},
  \label{eq:Schorodinger_equation_for_evolution}
\end{equation}
where $r$ is a real-valued parameter that describes how much the quantum state is changed. In contrast to the Schr\"odinger equation for a harmonic oscillator in Eq. (\ref{eq:Schrodinger_eq_for_quantum_state}), we don't limit the Hamiltonian $\hat H$ to be that of a harmonic oscillator. Instead, by choosing suitable Helmitian operator $\hat H$, we can modify the quantum state in various ways as shown in the following sections.

Eq. (\ref{eq:Schorodinger_equation_for_evolution}) can be solved as follows:
For an infinitesimal $\Delta r$, 
\begin{equation}
  \frac{i\hbar}{\Delta r} (\ket{\psi(\Delta r)} - \ket{\psi(0)}) = \hat H \ket{\psi(0)},
\end{equation}
\begin{equation}
  \ket{\psi(\Delta r)} = \left(1 - \frac{\hat H\Delta r}{i\hbar}\right) \ket{\psi(0)}.
\end{equation}
Here, we define $r = N\Delta r$ and take the limit of $N \to \infty$,
\begin{equation}
\begin{aligned}
  \ket{\psi(r)} &= \left(1 - \frac{\hat H\Delta r}{i\hbar}\right)^N \ket{\psi(0)}\\
  &\to \exp\left(\frac{\hat Hr}{i\hbar}\right)\ket {\psi(0)} \equiv \hat U(r)\ket {\psi(0)}.
\end{aligned}
\end{equation}
Since $\hat U(r) = \exp(\hat Hr/i\hbar)$ changes the quantum state according to the Hamiltonian, it is called \textbf{time-evolution operator}\index{time-evolution operator}. When $\hat H$ is the Hamiltonian of a harmonic oscillator, the time-evolution operator can change the quantum state so that the quantum state can evolve over time. In quantum optics, we use other type of Hamiltonian to change the quantum state. 

As we saw in Eq. (\ref{eq:diagonalization_of_exp_Hermitian}), since $\hat H$ is Hermitian, $\hat H$ can be diagonalized as $\hat H = \hat U_H\hat D\hat U_H^\dagger$. Here, $\hat U_H$ is the unitary operator composed of the eigenvectors of $\hat H$, and $\hat D$ is the diagonal matrix composed of the corresponding eigenvalues. Using them, we get
\begin{equation}
  \ket{\psi(r)} = \hat U_H\exp \left( \frac{\hat Dr}{i\hbar} \right)\hat U_H^\dagger \ket{\psi(0)}.
  \label{eq:diagonalized_evolution_operator}
\end{equation}
Eq. (\ref{eq:diagonalized_evolution_operator}) shows that the time-evolution operator applies the phase shift that is proportional to $d$, where $d$ is the eigenvalue of the Hamiltonian, depending on the eigenstate of the Hamiltonian. Since the eigenvalues of Hermitian operator is real, time-evolution operator does not change the norm of the quantum state, and therefore it is unitary.

\section{Heisenberg picture}
In the Heisenberg picture, instead of changing the quantum state, we change the operator of observable. This is possible because of the following reason: As we see in the previous chapter, the quantum-mechanical measurement involves the projection into eigenvectors of observable(s), and takes the weighted average of eigenvalues. Therefore, by changing the eigenvectors of the observable, we can calculate the same thing as the evolution of quantum state. In particular, we calculate the change of annihilation operator and creation operator to see what happens in the evolution of quantum state.

Specifically, the expectation value of an observable $\hat A$ for a quantum state $\ket{\psi(r)} = \hat U(r)\ket{\psi(0)}$ is given by
\begin{equation}
\begin{aligned}
  \braket{\psi(r)|\hat A|\psi(r)} &= \braket{\psi(0)|\hat U^\dagger(r)\hat A\hat U(r)|\psi(0)}\\
  &\equiv \braket{\psi(0)|\hat A_H(r)|\psi(0)}.
\end{aligned}
\end{equation}
where $\hat A_H(r) = \hat U^\dagger(r) \hat A \hat U(r)$. You can see that the eigenvector of the observable is changed by the unitary operator $\hat U(r)$.

The evolution of $\hat A_H(r)$ can be expressed by the Heisenberg equation of motion, which is derived as
\begin{equation}
  \begin{aligned}
  	\frac{d}{dr}\hat A_H(r) &= \frac{d}{dr}\left(\hat U^\dagger(r)\hat A\hat U(r)\right)\\
  	&= \left( \frac{d}{dr}\hat U^\dagger(r) \right)\hat A\hat U(r) + \hat U^\dagger(r)\hat A\frac{d}{dr}\hat U(r)\\
  	&= -\frac{\hat H }{i\hbar}\hat U^\dagger(r) \hat A\hat U(r) + \hat U^\dagger(r) \hat A \hat U(r)\frac{\hat H }{i\hbar}\\
  	&= \frac{i}{\hbar}(\hat H\hat A_H - \hat A_H \hat H) = \frac{i}{\hbar}[\hat H, \hat A_H]
  	\label{eq:Heisenberg_eq_motion}
  \end{aligned}
\end{equation}

\section{Time evolution}
\subsubsection{Schr\"odinger picture}
For calculating time evolution, we use the Hamiltonian of a harmonic oscillator $\hat H = \hbar\omega(\hat a^\dagger\hat a + 1/2) = \hbar\omega (\hat n + 1/2)$. The time evolution operator becomes
\begin{equation}
  \hat U_T(t) = e^{-i\omega (\hat n + 1/2)t},
\end{equation}
which indicates that $\hat U(t)$ applies different phase shift for different energy eigenstate $\ket n$, i.e.,
\begin{equation}
  \ket {\psi(t)} = \sum_{n = 0}^{\infty}e^{-i\omega(n+1/2)t}\ket n \braket{n|\psi(0)},
\end{equation}
which is the same as Eq. (\ref{eq:Schrodinger_eq_solution}).

\subsubsection{Heisenberg picture}
Here we calculate how annihilation operator and creation operator of a harmonic oscillator change along time in the Heisenberg picture. By substituting the Hamiltonian of the harmonic oscillator and $\hat A_H(t) = \hat a(t)$ to Eq. (\ref{eq:Heisenberg_eq_motion}), we obtain
\begin{equation}
\begin{aligned}
  \frac{d}{dt}\hat a(t) &= \frac{i}{\hbar}[\hbar\omega(\hat a^\dagger(t)\hat a(t) + 1/2), \hat a(t)]\\
  &= i\omega \left(\hat a^\dagger(t)\hat a(t)\hat a(t)-\hat a(t)\hat a^\dagger(t)\hat a(t)\right)\\
  &= i\omega \left(\hat a^\dagger(t)\hat a(t)\hat a(t)-(\hat a^\dagger(t)\hat a(t)+1)\hat a(t)\right)\\
  &= -i\omega \hat a(t)
\end{aligned}
\end{equation}
Therefore, 
\begin{equation}
  \hat a(t) = \hat a(0)e^{-i\omega t}.
  \label{eq:time_evolution_of_annihilation_operator}
\end{equation}
This is the quantum version of the time evolution of classical harmonic oscillator in Eq. (\ref{eq:normalized_complex_amplitude}). Similarly, we get
\begin{equation}
  \hat a^\dagger(t) = \hat a^\dagger(0)e^{i\omega t},
  \label{eq:time_evolution_of_creation_operator}
\end{equation}
\begin{equation}
\begin{aligned}
  \hat x(t) &= (\hat a(t) + \hat a^\dagger(t))/2
  = (\hat a(0)e^{-i\omega t} + \hat a^\dagger(0)e^{i\omega t})/2\\
  &= \hat x(0) \cos\omega t + \hat p(0) \sin \omega t,
  \label{eq:evolution_of_x}
\end{aligned}
\end{equation}
\begin{equation}
\begin{aligned}
  \hat p(t) &= (\hat a(t) - \hat a^\dagger(t))/2i
  = (\hat a(0)e^{-i\omega t} - \hat a^\dagger(0)e^{i\omega t})/2i\\
  &= -\hat x(0) \sin\omega t + \hat p(0) \cos \omega t.
\end{aligned}
\end{equation}
The meaning of Eq. (\ref{eq:evolution_of_x}) is clear. The measurement of $\hat x$ at time $t$ is equivalent to the measurement of linear combination of $\hat x(0)$ and $\hat p(0)$. This is a natural consequence of time evolution of a harmonic oscillator.

\subsubsection{Time evolution as Fourier transform}
It is noteworthy that, in the $x$ representation of a quantum state of a quantum harmonic oscillator, time evolution for 1/4 period is equivalent to inverse Fourier transform. This is because the time evolution corresponds to the rotation in phase space ($x$-$p$ space), and the wavefunction in $p$ is the Fourier transform of the wavefunction in $x$. 

Using
\begin{equation}
  \hat U_T(\pi/2\omega) = e^{-i(\pi/2)(\hat n + 1/2)} = e^{-i\pi/4}e^{-i\pi\hat n/2},
\end{equation}
we get

\begin{equation}
  \begin{aligned}
  	\braket{x|\hat U(\pi/2\omega)|n} &= e^{-i(n+1/2)\omega t}\bra x \frac{(\hat a^\dagger)^n}{\sqrt{n!}}\ket{0}\\
  	&= e^{-i\pi/2}\frac{(-i)^n}{\sqrt{n!}}\left(x - \frac 1 2 \frac{d}{dx} \right)^n\left(\frac 2 \pi \right)^{1/4}e^{-x^2}.\\
  \end{aligned}
\end{equation}
From Eq. (\ref{eq:creation_in_p}),
\begin{equation}
  \begin{aligned}
  	\braket{p|n} &= \bra p \frac{(\hat a^\dagger)^n}{\sqrt{n!}}\ket{0}\\
  	&= \frac{(-i)^n}{\sqrt{n!}}\left(p - \frac 1 2 \frac{d}{dp} \right)^n\left(\frac 2 \pi \right)^{1/4}e^{-p^2}.\\
  \end{aligned}
\end{equation}
Therefore,
\begin{equation}
  \begin{aligned}
  	\braket{x|\hat U(\pi/2\omega)|\psi} = \sum_{n=0}^\infty e^{-i\pi/2}\frac{(-i)^n}{\sqrt{n!}}\left(x - \frac 1 2 \frac{d}{dx} \right)^n\left(\frac 2 \pi \right)^{1/4}e^{-x^2}\braket{n|\psi}\\
  	\label{eq:time-evolved_wavefunction}
  \end{aligned}
\end{equation}
\begin{equation}
  \begin{aligned}
  	\braket{p|\psi} &= \sum_{n=0}^\infty \bra p \frac{(\hat a^\dagger)^n}{\sqrt{n!}}\ket{0}\braket{n|\psi}\\
  	&= \sum_{n=0}^\infty\frac{(-i)^n}{\sqrt{n!}}\left(p - \frac 1 2 \frac{d}{dp} \right)^n\left(\frac 2 \pi \right)^{1/4}e^{-p^2}\braket{n|\psi}.\\
  	\label{eq:time-evolved_wavefunction_p}
  \end{aligned}
\end{equation}
We can see that Eq. (\ref{eq:time-evolved_wavefunction}) and Eq. (\ref{eq:time-evolved_wavefunction_p}) are in the same form. Since $\tilde \psi(p) = \braket{p|\psi}$ is the Fourier transform of $\psi(x)=\braket{x|\psi}$, we can see that the time evolution over one fourth period is equivalent to Fourier transform and an additional phase shift of $e^{-i\pi/4}$. This means that, $\psi(x)$ after 1/4 period is almost the same as $\tilde \psi(p)$. This is analogous to classical harmonic oscillator where the complex amplitude rotates in the phase space and exchange the position and momentum energies. In this way, time evolution of quantum harmonic oscillator conducts Fourier transform.

\begin{comment}
\begin{equation}
  \begin{aligned}
	\braket{p|n} = \int\braket{p|x}\braket{x|n}	dx
  \end{aligned}
\end{equation}
\end{comment}

\section{Displacement}
Here we introduce the displacement operator given by
\begin{equation}
  \hat U_D(\alpha) = \exp(\alpha \hat a^\dagger - \alpha^* \hat a),
\end{equation}
where $\alpha$ is a complex number. Denoting the real part and the imaginary part of $\alpha$ as $x_0$ and $p_0$ such that $\alpha = x_0 + ip_0$, we will see that the displacement operator shifts the wavefunction $\phi(x)$ by $x_0$ and the momentum wavefunction $\tilde \phi(p)$ by $p_0$. 

By applying $\hat U_D(\alpha)$ to $\ket 0$, we can displace the complex amplitude of vacuum by $\alpha$ to obtain $\ket \alpha = \hat U_D(\alpha)\ket 0$. This is called coherent state, whose time evolution is sinusoidal motion. The coherent state is a quantum-mechanical description of classical wave. The property of coherent state will be described later.

Here we explain the function of displacement operator first in Heisenberg picture and Schro\"dinger picture. The former is simpler than the latter, while the latter is more intuitive.

\subsubsection{Heisenberg picture}
To see how the displacement operator works, we introduce the displacement Hamiltonian defined by
\begin{equation}
  \hat H_D = i\hbar (\alpha \hat a^\dagger - \alpha^* \hat a).
\end{equation}
Using this, the time-evolution operator is given by
\begin{equation}
\hat U(t) = \exp\left(\frac{t}{i\hbar}\hat H_D\right) = \exp\left\{t(\alpha\hat a^\dagger - \alpha^* \hat a)\right\},
\end{equation}
that is, $\hat U(1) = \hat U_D(\alpha)$. From the Heisenberg equation of motion, we get
\begin{equation}
 \frac{d}{dt}\hat a = \frac i \hbar[\hat H_D, \hat a] = -\alpha [\hat a^\dagger, \hat a] = \alpha,
\end{equation}
\begin{equation}
  \frac{d}{dt}\hat a^\dagger = \frac i \hbar [\hat H_D, \hat a^\dagger] = \alpha^*[\hat a, \hat a^\dagger] = \alpha^*.
\end{equation}
Therefore,
\begin{equation}
  \hat a(1) = \hat a(0) + \alpha.
  \label{eq:displaced_annihilation}
\end{equation}
\begin{equation}
  \hat a^\dagger(1) = \hat a^\dagger(0) + \alpha ^*.
  \label{eq:displaced_creation}
\end{equation}
Eq. (\ref{eq:displaced_annihilation}) shows that $\hat a = \hat x + i\hat p$ changes by $\alpha = x_0 + p_0$. This is the result of displacement operator in the Heisenberg picture.

\subsubsection{Schr\"odinger picture}
Here we explain why $\hat U_D(\alpha)$ can apply the displacement to the wavefunction. Since the displacement Hamiltonian $\hat H(D)$ is Hermitian, we can calculate its eigenvectors and eigenvalues. $U_D(\alpha)$ applies the phase shift according to the eigenvalues. We can calculate the eigenvectors but it's quite lengthy.

Instead, let's see what happens when $p_0 = 0$, i.e., $\alpha = x_0$. In this case,
 \begin{equation}
  \hat H_D = i\hbar x_0(\hat a^\dagger - \hat a) = i\hbar x_0\left\{\hat x-i\hat p - (\hat x +i\hat p)\right\} = 2\hbar x_0 \hat p.
\end{equation}
and $\hat U(t) = \exp(-2ix_0t\hat p)$. Therefore,
\begin{equation}
  \hat U(t)\psi(x) = \exp(-2ix_0t\hat p)\psi(x) = \exp\left(x_0t\frac{\partial }{\partial x}\right) \psi(x),
\end{equation}
which leads to
\begin{equation}
  \frac{\partial}{\partial t}\hat U(t)\psi(x) = x_0\frac{\partial}{\partial x}\hat U(t)\psi(x).
\end{equation}
Since this is a wave equation, you can see that $\hat U(t)\psi(x)$ propagates at a speed of $x_0$, and hence $\hat U(1)$ shifts $\psi(x)$ by $x_0$.

Another view is to use Fourier transform. $\hat U(1) = \exp(-2ix_0\hat p)$ applies phase shift to the eigenvector of $p$, and the amount of phase shift is proportional to $-2x_0p$. This is equivalent to applying phase shift to momentum wavefunction, i.e., $\tilde\psi(p) \to \tilde\psi(p)e^{-2ix_0p}$. The resultant wavefunction $\psi'(x)$ is given by the inverse Fourier transform:
\begin{equation}
  \psi'(x) = \int \tilde \psi(p)e^{-2ix_0p}e^{2ixp}dp = \psi(x-x_0).
\end{equation}
In this way, displacement in $x$ can be viewed as the shift theorem of Fourier transform. 

The displacement operator is an expanded version of shift theorem so that not only $x$ but also $p$ can be displaced by $\alpha = x_0 + ip_0$. To see what happens, we divide this process into three: 1. Clockwise rotation by $\theta = \arg \alpha$ in $x$-$p$ plane. 2. Shift in $x$ position by $|\alpha|$. 3. Counterclockwise rotation by $\theta$ in $x$-$p$ plane. Since time evolution is clockwise rotation in $x$-$p$ plane, the whole process can be described as
\begin{equation}
\begin{aligned}
  \hat U_T(-\theta/\omega)\exp(-2i|\alpha|\hat p)\hat U_T(\theta/\omega) &= \hat U_T^\dagger(\theta/\omega)\exp\left\{|\alpha|(\hat a^\dagger - \hat a)\right\}\hat U_T(\theta/\omega)\\
  &= \exp\left\{ |\alpha|(\hat a^\dagger e^{i\omega t} - \hat ae^{-i\omega t})\right\}\\
  &= \exp(\alpha \hat a^\dagger - \alpha^* \hat a).
\end{aligned}
\end{equation}
From the 1st line to the 2nd line, we converted $\hat a$ and $\hat a^\dagger$ to $\hat ae^{-i\omega t}$ and $\hat a^\dagger e^{i\omega t}$ using Eq. (\ref{eq:time_evolution_of_annihilation_operator}) and Eq. (\ref{eq:time_evolution_of_creation_operator}). This is possible because the RHS in the 1st line is the same form of the operator modified by the Heisenberg equation of motion.

\section{Mode mixing}
There are various ways to see the linear combination of several modes. In optics, there are various situations where the linear combination of light is obtained such as: 
\begin{itemize}
	\item A beam splitter combines two beams to see the sum and difference of two light waves \textcolor{red}{(Fig. ??)}.
	\item Waveplates can rotate and change the polarization to mix the horizontally polarized light and the vertically polarized light.
	\item  
\end{itemize}



\section{Single-mode squeezing}

\section{Two-mode squeezing}

\section{Summary}




\chapter{Quantization of light}
\section{Mode decomposition of electromagnetic waves}
\subsection{Time-frequency mode}
\subsection{Spatial mode}
\subsection{Polarization}
\section{Operator notation of electromagnetic waves}
\section{Summary}

\chapter{Representative quantum states}
\section{Number states}
\section{Superposition states}
\section{Coherent states}
\section{Squeezed states}
\section{Two-mode squeezed states}
\subsection{EPR state}
\section{Summary}

\chapter{Control of quantum states of light}
\section{Mode mixing}
\subsection{Beamsplitter}
\subsection{Waveplates}
\subsection{Optical loss}
\subsection{Time-frequency modes}
\section{Parametric amplification}
\subsection{Squeezing}
\subsection{Spontaneous parametric down conversion}
\subsection{Optical amplification}
\subsection{Raman scattering}
\section{Summary}

\chapter{Quantum-optical measurement}
\section{Direct detection}
\section{Homodyne detection}
\section{Heterodyne detection}
\section{Preamplification}
\section{Quantum teleportation}
\section{Summary}

\appendix
\chapter{Appendix}
\section{Bra-ket notation}
\section{Creation and annihilation operators}
\section{Pure states and mixed states}
\section{Wigner function}

\begin{equation}
\begin{aligned}
  \sum_{k=1}^\infty \frac 1 {2^k} &= \frac 1 {2^1} + \frac 1 {2^2} + \frac 1 {2^3} + \dots \\
  &= \frac{1}{2} + \frac{1}{4} + \frac{1}{8} + \dots \\
  &= \frac{\frac 1 2}{1-\frac 1 2} =  1
\end{aligned}
\end{equation}

This is a simple calculation \cite{adams1995hitchhiker}.

\printindex

\bibliographystyle{plain}
\bibliography{references}
\end{document}
